Fast and Scalable Mixed Precision Euclidean Distance Calculations Using GPU Tensor Cores

### 2025-05-01 Paper/main.tex
Added paper from ICPP2025 submission.

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf'.tex)
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf, screen, review]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{none} 
%\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference['ICS 2025']{The 39th ACM International Conference on
  Supercomputing}{June 9--11, 2025}{Salt Lake City, USA}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
\acmISBN{000-0-0000-0000-0/00/00}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%custom commands

\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{multirow}
\usepackage[noend]{algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[dvipsnames]{xcolor}
\usepackage{forest, tikz}
\usepackage{enumitem}
\usepackage{sidecap}
\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{pgfplots}
\usepackage{tikzscale}
% \usepackage[disable]{todonotes} %disable
\usepackage[]{todonotes} %enable
\usepackage{listings}
\lstset{
language=C++,
basicstyle=\small,
showstringspaces=false,
}

@@

\usepackage{tikz}
\usepackage{xcolor}
\usepackage{xparse}
\usetikzlibrary{calc}
\usetikzlibrary{matrix}
\usetikzlibrary{arrows.meta}
\usepackage{xparse}

\usepackage{xspace}    
\usepackage{tabularx}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,fill,inner sep=1.3pt] (char) {\footnotesize\textcolor{white}{#1}};}}

\newcommand{\mike}[1]{\todo[color=red, inline]{Mike: #1}}
\newcommand{\brian}[1]{\todo[color=yellow,inline]{Brian: #1}}

@@

%placeholder
\def\ouralg{\textsc{MPTC-Join}\xspace}
\def\tedjoin{\textsc{TED-Join}\xspace}
\def\tedjoinbrute{\textsc{TED-Join-Brute}\xspace}
\def\tedjoinindex{\textsc{TED-Join-Index}\xspace}
\def\gdsjoin{\textsc{GDS-Join}\xspace}

%datasets
\def\tinyds{\textit{Tiny5M}\xspace}
\def\sift{\textit{Sift10M}\xspace}
\def\gist{\textit{Gist1M}\xspace}
\def\cifar{\textit{Cifar60K}\xspace}
\def\msd{\textit{MSD}\xspace} 
\def\bigcross{\textit{Bigcross}\xspace} 
\def\susy{\textit{SuSy}\xspace} 
\def\higgs{\textit{Higgs}\xspace} 
\def\wave{\textit{Wave}\xspace} 
\def\uniform{\textit{Uniform}\xspace} 
\def\expo{\textit{Expo}\xspace}

@@

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}
\settopmatter{printfolios=true}
\settopmatter{printacmref=false}
%\pagestyle{plain}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

\input{figures/diagramResources.tex}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Mixed Precision Euclidean Distance Calculations using GPU Tensor Cores and Application to Similarity Searches}
\subtitle{\normalsize{ICS 2025 Submission
    \textbf{\#NaN} -- Confidential Draft -- Do NOT Distribute!!}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Uncomment upon acceptance
% \author{Brian Curless}
% \affiliation{%
%   \institution{School of Informatics, Computing, \& Cyber Systems\\Northern Arizona University}
%   \city{Flagstaff}
%   \state{Arizona}
%   \country{USA}
% }
% \email{bc2497@nau.edu}

% \author{Michael Gowanlock}
% \affiliation{%
%   \institution{School of Informatics, Computing, \& Cyber Systems\\Northern Arizona University}
%   \city{Flagstaff}
%   \state{Arizona}
%   \country{USA}
% }
% \email{Michael.Gowanlock@nau.edu}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Tensor cores (TCs) have been shown to offer speedups over typical CUDA core-based distance similarity self-join problems. However, past work has only investigated using FP64 operands for computation. Modern TCs have the ability to operate on lower precision floating point operands as well. Ampere architecture GPUs can even compute matrix multiplications with half-precision inputs (FP16), while accumulating the results of the computation in single precision (FP32). This mixed-precision configuration offers the highest computational throughput out of all floating-point Tensor core configurations available on the Ampere architecture. To achieve the maximum throughput of 312 teraflops, one needs to carefully design their algorithm for significant hierarchical reuse of data, and maximize memory utilization at every level (global, shared, registers). Previous work utilizes a search index data structure to pare the search space down and reduce the number of computations required. In order to maximize the use of GPU hardware, the problem lends itself to a cohesive brute force strategy. This paper compares our cohesive brute force, mixed-precision Tensor core distance similarity self-join algorithm to a state of the art double precision (FP64) Tensor core algorithm, and a state of the art single precision (FP32) cuda core algorithm. We find that on high-dimensional datasets, the brute force approach outperforms the state of the art by a factor of 4-16, depending on the dataset. Additionally, we quantify the accuracy loss of our reduced precision algorithm to be <0.04\%.
\end{abstract}

@@

\mike{Notes on experiments GDS-Join: batch size for GDS-Join is 200,000,000. Need to remove the time to perform batch estimation.}
\mike{Brian: make sure to remove the time for the batch estimator for TED-Join. Run TED-Join with a batch size of 200,000,000 to be consistent with GDS-Join.}
\mike{Experiments: need to add the increased shared memory allocation for TED-Join}
\brian{I added this to the Tedjoin section}

\mike{From Brian's preliminary results it looks like the time for his algorithm may be less than the time needed to index with GDS-Join!!}
\section{Introduction}\label{sec:intro}
\mike{Define TCs at the beginning}

Euclidean distance calculations are fundamental to numerous applications. For example, in the area of data analytics, they are employed for similarity searches~\cite{gowanlock2023optimization}, outlier detection~\cite{zimek2012survey}, k-nearest neighbor searches~\cite{Samet2008}, and clustering~\cite{bottesch2016speeding}, among other tasks. The Euclidean distance calculation is often a major bottleneck in an application, and given the numerous algorithms that rely on Euclidean distance subroutines, improving the performance of this fundamental operation may have a significant impact on numerous application domains.

In this paper, we propose \ouralg, which is a highly optimized GPU tensor core (TC) algorithm that computes Euclidean distance calculations and which is designed for moderate to high dimensional datasets. We consider comparing all points in an input dataset to each other, which has a quadratic time complexity, and only return those points that are within a search distance, $\epsilon$, of each other. This is a common application scenario, as it is assumed or employed as an optimization technique in many of the data analytics applications described above.

Peak throughput, defined here as the maximum number of Tera Floating Point Operations Per Second (TFLOPS) yields an upper bound on performance, as it is the maximum throughput achievable based on hardware constraints. The Nvidia A100 GPU that we use in our evaluation has a peak throughput using FP32 (single precision floats) on CUDA cores of 19.5 TFLOPS, whereas TCs can achieve a throughput of 312 TFLOPS using mixed precision floats (FP16 operands and FP32 accumulation)~\cite{A100}. Given that the peak throughput of TCs outlined above is a factor 16 higher than CUDA cores, it is clear that if they can be efficiently exploited to compute Euclidean distances, there is immense potential for significant performance improvements over the state-of-the-art methods that only employ CUDA cores.

Based on our investigation, we find that there are three major challenges that limit the peak performance achievable on GPU TCs for Euclidean distance calculations.

% [label={{\bf D\arabic*}}]
\begin{enumerate}
\item \bfseries{Memory Bottlenecks --} A major challenge that we identify is that the TCs will be underutilized if the data is not efficiently transferred between each level of the GPU's memory hierarchy, i.e., from global memory, to shared memory, and then into registers for computation on TCs. We find that the coordinates of data points need to be stored and accessed in a manner that maximizes bandwidth utilization and eliminates shared memory bank conflicts during loading and storing.
\item \bfseries{Insufficient Programming Flexibility using the Warp Matrix Multiply Accumulate (WMMA) API --} The WMMA API~\cite{wmmaNvidia} provides programmatic access to TCs where threads in a warp cooperate to execute MMA operations on matrix fragments. These fragments are arrays of registers that represent a subset of a larger matrix. The WMMA API has rigid load/store options that yield the bank conflict problem outlined above, and hide important details about how data flows into TCs.

%\mike{Brian: See (3) below. It seems to me that GeMM has an $O(N^3)$ time complexity, whereas Euclidean distance calculations have $O(|D|^2)$. Therefore one would expect that GeMM would be able to better utilize TCs. But, in our paper, we really have a hidden factor $d$, where the time complexity is $O(d\cdot|D|^2)$. But, in practice $d\ll|D|$, so we ignore $d$. But I know you've said that Euclidean distance calculations take the same work as matrix multiply, so I think I'm missing something here.}
%\brian{I think everything you wrote is correct, I think it must be something small that you are missing. GeMM requires $m*n*k$ computations, and distance requires $d*|D|^2$. In this case though, $d = k$ and $m = n = |D|$. So they are identical computations, with the euclidian distance having the sum of the square values added in at the end (which we can ignore when considering complexity). Also, I adjusted (3), hopefully that helps clear things up.}
\item \bfseries{Insufficient Data Reuse --} General Matrix Multiplication, one of the primary applications of TCs, has a high compute to memory access ratio because of its data reuse, and so the peak throughput achievable on TCs is very high. The highest time complexity component of computing the Euclidean distance between all pairs of points in a dataset can be formulated as a GEMM. In order to ensure the TCs are not idle and waiting for registers to be loaded with data elements, the data reuse has to be carefully designed at all levels including DRAM, L2 Cache, Shared Memory, and registers.

@@

In this paper, we address the challenges above and make the following contributions.

\begin{enumerate}
\item When copying matrix data into the streaming multiprocessors, we optimize the entire data path from global memory into registers. We employ inline Parallel Thread Execution (PTX) instructions that provide the flexibility of non-standard memory access patterns, such that load/store instructions do not cause shared memory bank conflicts.
\item We design a two stage pipeline that includes asynchronous memory copies for loading data from global memory into shared memory and then into registers, such that each warp overlaps TC computation with asynchronous loads. This reduces the duration of memory stalls from global memory, and memory pressure on limited shared memory resources.
\item We optimize storing data, such that each CUDA block stores multiple "k-slices" of matrix data in shared memory, and each warp stores multiple matrix fragments in registers that create data reuse. We also dispatch the work in a specific order to achieve high L2 cache reuse that obviates global memory accesses. 
\item We show that \ouralg can achieve up to XXX\% of peak mixed precision TC performance. We compare \ouralg, to both TC and CUDA core implementations in the literature and demonstrate that it is superior in most experimental scenarios.
\item We demonstrate that when compared with an algorithm using double precision operands, a mixed precision algorithm can produce highly accurate results on real world datasets.
\end{enumerate}

The paper is outlined as follows. Section~\ref{sec:background} outlines background material, including definitions, the operation of tensor cores, and related work. Section~\ref{sec:our_algorithm} outlines our proposed Euclidean distance algorithm using TCs. Section~\ref{sec:evaluation} evaluates our algorithm and compares it to the state-of-the-art. Finally, Section~\ref{sec:conclusion} concludes the work and outlines future research directions.

@@

\subsection{Definitions and Problem Statement}\label{sec:problem_statement}
\noindent\textbf{Definition of a dataset and data points --} Let $D$ be a dataset containing $|D|$ points, where each point is defined as $p_i\in D$, and $i=1, 2, \ldots, |D|$. Each point, $p_i$, is defined by coordinates in $d$ dimensions. We define the coordinates for point $p_i$ as $x_j$, where $j=1, 2, \ldots, d$, and therefore, $x_j=\{x_1, x_2, \ldots, x_d\}$.

\noindent\textbf{Definition of the Euclidean Distance --} In what follows, we outline the definition of the Euclidean distance, where we first describe the standard definition and then describe the definition that can be computed on tensor cores. Let $p\in D$ and $q \in D$ be two points in the dataset, $D$, where $p_j$ and $q_j$ refer to the $j^{th}$ coordinates of points $p$ and $q$, respectively. The Euclidean distance between $p$ and $q$ is typically defined as follows:

\begin{equation}
dist(p, q) = \sqrt{\sum^{d}_{j=1}(p_j - q_j)^2}.
\label{eqn:euclidean_distance}
\end{equation}

While this definition is impossible to compute on TCs, a simple expansion yields a form that can be easily computed on TCs: ~\cite{gallet2022leveraging}:

\begin{equation}
dist(p, q) = \sqrt{\sum^{d}_{j=1}p_j^2 - 2p_jq_j + q_j^2}.
\label{eqn:euclidean_distance_TCs}
\end{equation}

\noindent\textbf{Definition of queries and the self-join --} In this paper we compare all points $p_i \in D$ to each other, which is a high throughput operation that is well-suited to the GPU. In database terminology, this is referred to as a self-join, where conceptually $D$ can be considered a database table, and it is joined on itself. A straightforward implementation of this operation requires three nested for loops.

\subsection{Two Scenarios for Euclidean Distance Calculations}
There are numerous applications that require the calculation of Euclidean distances (Section~\ref{sec:intro}). We describe three common application scenarios below.

\subsubsection{Scenario 1: Brute Force Distance Similarity Searches}\label{sec:scenario1_brute_with_epsilon}
In this scenario, all points are searched in a dataset $D$ (containing $|D|$ points), and it returns all result pairs where the distance is $\leq\epsilon$, i.e., the self-join described in Section~\ref{sec:problem_statement}. This has a time complexity of $O(|D|^2)$.

\subsubsection{Scenario 2: Index-Supported Distance Similarity Searches}\label{sec:scenario2_index}
To improve the performance of brute force similarity searches (described above), algorithms will typically first construct an indexing data structure that is used to prune the search between points that are sufficiently far from each other in the data space. Then, when a given query point is compared to all other points in the dataset, it only needs to consider a subset of the $|D|$ points, thus reducing the total number of distance comparisons that are required of a brute force search. While our proposed algorithm, \ouralg, does not use an index, we compare it to state-of-the-art GPU implementations do use one.

@@

In this section, we discuss the operation of TCs and how they are programmed on Nvidia GPUs. Tensor Cores can operate on many different kinds of floating point data including FP64, FP32, and FP16-FP32 mixed precision operands. There are several libraries that directly employ TCs for linear algebra operations, including cuBLAS~\cite{cublas} and CUTLASS~\cite{cutlass}. These libraries are suitable when their subroutines can be directly embedded into an algorithm. However, they are unsuitable for our purposes, as they do not offer sufficient granularity and control. There are two other ways to use TCs that provide more flexibility, the WMMA API, and inline PTX instructions.~\cite{wmmaNvidia}. 
\brian{This citation might need PTX instrucitons added too.}

>>>>

%%
%% This is file `sample-sigconf-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,authordraft')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf-authordraft.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf,anonymous, review]{acmart}
%\documentclass[sigconf,authorversion]{acmart}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%% End of template, start of our custom Preamble here:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%custom commands
% For importing pgf plots
\usepackage[utf8]{inputenc}
\usepackage{colortbl}
\usepackage{amsmath,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{multirow}
\usepackage[noend]{algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[dvipsnames]{xcolor}
\usepackage{forest, tikz}
\usepackage{enumitem}
\usepackage{sidecap}
\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\usepackage{tikzscale}
% \usepackage[disable]{todonotes} %disable
\usepackage[]{todonotes} %enable
\usepackage{listings}
\lstset{
language=C++,
basicstyle=\small,
showstringspaces=false,
}
\usepackage{bm}

@@

\usepackage{tikz}
\usepackage{xcolor}
\usepackage{xparse}
\usetikzlibrary{calc}
\usetikzlibrary{matrix}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{backgrounds}
\usepackage{xparse}

\usepackage{xspace}    
\usepackage{tabularx}

\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,fill,inner sep=1.3pt] (char) {\footnotesize\textcolor{white}{#1}};}}

\newcommand{\mike}[1]{\todo[color=red, inline]{Mike: #1}}
\newcommand{\brian}[1]{\todo[color=yellow,inline]{Brian: #1}}
%\renewcommand\brian[1]{}
%\renewcommand\mike[1]{}

@@

%placeholder
% \def\ouralg{\textsc{MPTC-Join}\xspace}
\def\ouralg{\textsc{FaSTED}\xspace}
\def\tedjoin{\textsc{TED-Join}\xspace}
\def\tedjoinbrute{\textsc{TED-Join-Brute}\xspace}
\def\tedjoinindex{\textsc{TED-Join-Index}\xspace}
\def\gdsjoin{\textsc{GDS-Join}\xspace}
\def\mistic{\textsc{MiSTIC}\xspace}

\def\ptxmma{PTX \texttt{mma}}

%datasets
\def\tinyds{\textit{Tiny5M}\xspace}
\def\sift{\textit{Sift10M}\xspace}
\def\gist{\textit{Gist1M}\xspace}
\def\cifar{\textit{Cifar60K}\xspace}
\def\msd{\textit{MSD}\xspace} 
\def\bigcross{\textit{Bigcross}\xspace} 
\def\susy{\textit{SuSy}\xspace} 
\def\higgs{\textit{Higgs}\xspace} 
\def\wave{\textit{Wave}\xspace} 
\def\uniform{\textit{Uniform}\xspace} 
\def\synth{\textit{Synth}\xspace}

\newcommand\kslice{$k$-slice\xspace}
\newcommand\kslices{$k$-slices\xspace}
\newcommand\eightd{$d{=}8$\xspace}
\newcommand\sixtd{$d{=}16$\xspace}
\newcommand\sixfourd{$d{=}64$\xspace}

\newcommand\regf{\text{rf}}
\newcommand\blockf{\text{bf}}
\newcommand\arf{\ensuremath{\bm{A}_{\regf}}}
\newcommand\prf{\ensuremath{\bm{P}_{\regf}}}
\newcommand\qrf{\ensuremath{\bm{Q}_{\regf}}}
\newcommand\pbf{\ensuremath{\bm{P}_{\blockf}}}
\newcommand\qbf{\ensuremath{\bm{Q}_{\blockf}}}

\newcommand\pointRange[2]{%
    \def\firstPoint{#1}
    \def\lastPoint{#2}

    \ensuremath{\{p_{\firstPoint}, \dotsc, p_{\lastPoint\}}}}

\newcommand\dimRange[2]{%
    \def\firstDim{#1}
    \def\lastDim{#2}

    \ensuremath{k \in \{\firstDim,\dotsc, \lastDim\}}}

\definecolor{OliveGreen}{cmyk}{0.64,0,0.95,0.40}
\definecolor{CadetBlue}{cmyk}{0.62,0.57,0.23,0}
\definecolor{lightlightgray}{gray}{0.93}
\lstset{
language=C++,                          % Code langugage
basicstyle=\ttfamily,                   % Code font, Examples: \footnotesize, \ttfamily
keywordstyle=\color{OliveGreen},        % Keywords font ('*' = uppercase)
commentstyle=\color{gray},              % Comments font
numbers=left,                           % Line nums position
numberstyle=\tiny,                      % Line-numbers fonts
stepnumber=1,                           % Step between two line-numbers
numbersep=5pt,                          % How far are line-numbers from code
backgroundcolor=\color{lightlightgray}, % Choose background color
frame=none,                             % A frame around the code
tabsize=2,                              % Default tab size
captionpos=t,                           % Caption-position = bottom
breaklines=true,                        % Automatic line breaking?
breakatwhitespace=false,                % Automatic breaks only at whitespace?
showspaces=false,                       % Dont make spaces visible
showtabs=false,                         % Dont make tabls visible
columns=flexible,                       % Column format
morekeywords={__global__, __device__},  % CUDA specific keywords
}

\usepackage[most]{tcolorbox}
\definecolor{mGreen}{rgb}{0,0.6,0}
\definecolor{mGray}{rgb}{0.5,0.5,0.5}
\definecolor{mPurple}{rgb}{0.58,0,0.82}
\definecolor{backgroundColour}{rgb}{0.95,0.95,0.92}

%enable/disable tikz figures
\usepackage{comment}

%Comment/uncomment below to turn on/off tikz because of long compile times
%\excludecomment{tikz}
\includecomment{tikz}

%\excludecomment{todo}
%\includecomment{todo}

@@

%%
%% end of the preamble, start of the body of the document source.
\begin{document}
\begin{tikz}
\input{figures/diagramResources.tex}
\end{tikz}

%Potential names for the algorithm
% Fast and Scalable Tensor Euclidean Distance
% FaSTED

% Mixed Precision Tensor Euclidean Distance
% MPTED

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Fast and Scalable Mixed Precision Euclidean Distance Calculations Using GPU Tensor Cores}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Uncomment upon acceptance
\author{Brian Curless}
\affiliation{%
  \institution{School of Informatics, Computing, \& Cyber Systems\\Northern Arizona University}
  \city{Flagstaff}
  \state{Arizona}
  \country{USA}
}
\email{bc2497@nau.edu}

\author{Michael Gowanlock}
\affiliation{%
  \institution{School of Informatics, Computing, \& Cyber Systems\\Northern Arizona University}
  \city{Flagstaff}
  \state{Arizona}
  \country{USA}
}
\email{Michael.Gowanlock@nau.edu}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.

\renewcommand{\shortauthors}{Curless and Gowanlock}

\begin{abstract}
Modern GPUs are equipped with tensor cores (TCs) that are commonly used for matrix multiplication in artificial intelligence workloads. However, because they have high computational throughput, they can lead to significant performance gains in other algorithms if they can be successfully exploited. We examine using TCs to compute Euclidean distance calculations, which are used in many data analytics applications. Prior work has only investigated using 64~bit floating point (FP64) data for computation; however, TCs can operate on lower precision floating point data (i.e., 16 bit matrix multiplication and 32 bit accumulation), which we refer to as FP16-32. FP16-32 TC peak throughput is so high that TCs are easily starved of data.  We propose a Fast and Scalable Tensor core Euclidean Distance (FaSTED) algorithm. To achieve high computational throughput, we design  \ouralg for significant hierarchical reuse of data and maximize memory utilization at every level (global memory, shared memory, and registers).  We apply \ouralg to the application of similarity searches, which typically employ an indexing data structure to eliminate superfluous Euclidean distance calculations. We compare to the state-of-the-art (SOTA) TC Euclidean distance algorithm in the literature that employs FP64, as well as to two single precision (FP32) CUDA core algorithms that both employ an index. We find that across four real-world high-dimensional datasets spanning 128-960 dimensions, the mixed-precision brute force approach achieves a speedup over the SOTA algorithms of 2.5--51$\times$. We also quantify the accuracy loss of our mixed precision algorithm to be $<$0.03\% when compared to the FP64 baseline.

\end{abstract}

@@

\section{Introduction}\label{sec:intro}

Euclidean distance calculations are fundamental to numerous application domains (e.g., in the area of data analytics, they are employed for distance similarity searches~\cite{gowanlock2023optimization}, outlier detection~\cite{zimek2012survey}, k-nearest neighbor searches~\cite{Samet2008}, and clustering~\cite{bottesch2016speeding}). Given the numerous algorithms that rely on Euclidean distance subroutines, improving the performance of this fundamental operation can have a significant impact on a wide range of algorithms and applications.

In this paper, we present a Fast and Scalable Tensor core Euclidean Distance (\ouralg) algorithm, that is designed for moderate to high-dimensional datasets.\footnote{The code will be made publicly available after paper acceptance.} We consider comparing all points in an input dataset to each other, which has a quadratic time complexity, and only return those points that are within a search distance, $\epsilon$, of each other. This is a common application scenario, as it is assumed in many of the data analytics applications described above. \ouralg can be employed as a subroutine in algorithms and workloads requiring the large scale computation of Euclidean distances.

We define peak throughput as the maximum number of Tera Floating Point Operations Per Second (TFLOPS), which is the maximum throughput achievable based on TC hardware constraints. The Nvidia A100 GPU that we use in our evaluation has a peak FP32 throughput on CUDA cores of 19.5 TFLOPS, whereas TCs can achieve a throughput of 312 TFLOPS using FP16-32~\cite{A100}.\footnote{We use CUDA terminology throughout this paper.} Given that the peak throughput of TCs outlined above is a factor of 16 greater than CUDA cores, it is clear that if they can be efficiently exploited to compute Euclidean distances, there is potential for significant performance improvements over methods that employ CUDA cores.

Based on our investigation, we find that there are two major challenges that limit the performance achievable on GPU TCs for Euclidean distance calculations: 
\begin{enumerate}[wide=0pt]
\item \bfseries{Memory Bottlenecks --} A major challenge is efficiently transferring data between each level of the GPU's memory hierarchy, from global memory to shared memory to registers.

\item \bfseries{Insufficient Data Reuse --} Given the enormous disparity between FP16-32 TC throughput and global memory bandwidth, exploiting data reuse is critical. 
\end{enumerate}

This paper makes the following contributions:

\begin{enumerate}[wide=0pt]
\item When copying matrix data into streaming multiprocessors (SMs), we employ inline Parallel Thread Execution (PTX) instructions to design efficient memory access patterns.
\item We calculate the data reuse required to maximize performance and design warp- and block-level optimizations to meet these prerequisites. 
\item We design a two-stage pipeline with asynchronous memory copies for loading data from global into shared memory, such that each warp overlaps TC computation with asynchronous loads. %This reduces global memory stalls and register pressure.
%\item We dispatch work in a deterministic order to achieve high L2 cache reuse that obviates slow global memory accesses. 
\item We compare \ouralg to both TC and CUDA core state-of-the-art (SOTA) algorithms and demonstrate that it is superior in all experimental scenarios.

%We show that \ouralg can achieve up to 49\% of peak FP16-32 TC performance.

@@

The paper is outlined as follows: Section~\ref{sec:background} provides relevant background information on similarity searches and an exploration of related work; Section~\ref{sec:our_algorithm} describes how \ouralg functions; Section~\ref{sec:evaluation} evaluates \ouralg; and, finally, Section~\ref{sec:conclusion} reflects on the work's central contributions and future research directions.

@@

\subsection{Definitions and Problem Statement}\label{sec:problem_statement}
\noindent\textbf{Definition of a dataset and data points --} Let $D$ be a dataset containing $|D|$ points, where each point is defined as $p_i\in D$, and where $i{=}1, 2, \ldots, |D|$. Each point, $p_i$, is defined by coordinates in $d$ dimensions. We define the coordinates for point $p_i$ as $p_{i,k}$, where $k{=}1, 2, \ldots, d$.

\noindent\textbf{Definition of the Euclidean Distance --}  
Let $p_i\in D$ and $p_j \in D$ be two points in the dataset, $D$, where $p_{i,k}$ and $p_{j,k}$ refer to the $k^{th}$ coordinates of points $p_i$ and $p_j$, respectively. The Euclidean distance between $p_i$ and $p_j$ is typically defined as follows and an expansion yields a form that can be computed on TCs~\cite{gallet2022leveraging}:

{
\addtolength{\belowdisplayskip}{-1.0ex}
\addtolength{\belowdisplayshortskip}{-0.5ex}
\addtolength{\abovedisplayskip}{-2.0ex}
\addtolength{\abovedisplayshortskip}{-2.0ex}
% \begin{equation}
% dist(p_i, p_j) = \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2}= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_ip_j + p_{j,k}^2}.
% \label{eqn:euclidean_distance}
% \end{equation}
\begin{equation}
dist(p_i, p_j) = \sqrt{\sum^{d}_{k=1}(p_{i,k} - p_{j,k})^2}= \sqrt{\sum^{d}_{k=1}p_{i,k}^2 - 2p_{i,k}p_{j,k} + p_{j,k}^2}.
\label{eqn:euclidean_distance}
\end{equation}
}

\noindent\textbf{Definition of queries and the distance similarity self-join --} In this paper, we compare all points $p_i \in D$ to each other, which is a high throughput operation that is well-suited to the GPU. In database terminology, this is referred to as a self-join, where conceptually $D$ can be considered a database table that is joined with itself. This is a typical application scenario for the data analytics applications described in Section~\ref{sec:intro}.

\subsection{Scenarios for Euclidean Distance Calculations}
We describe two common application scenarios for the calculation of Euclidean distances below.

\subsubsection{Scenario 1: Brute Force Distance Similarity Searches}\label{sec:scenario1_brute_with_epsilon}
The distance is calculated between all permutations of points in a dataset $D$ (containing $|D|$ points), and only the pairs where the distance is $\leq\epsilon$ are returned. This has a time complexity of $O(|D|^2)$.

\subsubsection{Scenario 2: Index-Supported Distance Similarity Searches}\label{sec:scenario2_index}
To improve the performance of brute force similarity searches (described above), algorithms will typically first construct an indexing data structure. The index is used to prune the search between points that are sufficiently far from each other, decreasing the amount of comparisons required. %While our proposed algorithm, \ouralg, does not use an index, we compare it to GPU algorithms that use an index.

@@

In this section, we discuss the operation of TCs and how they are programmed on Nvidia GPUs. TCs can operate on different floating point data types including FP64, FP32, and FP16-32. There are several libraries that directly employ TCs for linear algebra operations, including cuBLAS and CUTLASS.\footnote{cuBLAS:  \url{https://developer.nvidia.com/cublas}; CUTLASS: \url{https://github.com/NVIDIA/cutlass}} These libraries are suitable when their subroutines can be directly embedded into an algorithm. However, they are unsuitable for our purposes, as they do not offer sufficient granularity or control. There are two other ways to use TCs that provide more flexibility: the WMMA API~\cite{wmmaNvidia}, and PTX instructions.\footnote{\url{https://docs.nvidia.com/cuda/archive/12.0.1/cuda-c-programming-guide/\#element-types-and-matrix-sizes}}

The WMMA API and PTX instructions both load data from shared memory into registers that are shared across an entire warp to compute small submatrices (called register fragments). However, the WMMA API is limited to larger matrix sizes (see Table~\ref{tab:WMMA_vs_PTX}), does not specify the register layout, and yields less control over memory addressing. In this paper, we use PTX instructions as they provide the most flexibility to achieve the highest throughput.

\begin{table}[]
